Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-09-03 manuscript.qmd
make risks dynamic, too
Also adds paragraph about LNLs I, V, and VII.

Related: #12

@Fig-model-risks (left panel) shows the predicted risk of occult disease in contralateral LNL II. The most important variable impacting the prediction for contralateral level II involvement in our model is the tumor's lateralization. A patient with a clearly lateralized early T-category tumor and a clinically N0 neck is predicted to have a 1-2% <!-- TODO: make this dynamic --> risk for occult disease in contralateral LNL II. For a patient with a clearly lateralized early T-category tumor but *with* mid-sagittal extension, the risk increases to almost 7%. <!-- TODO: make this dynamic -->

Also advanced T-category increases the risk of occult disease but plays a lesser role: Considering the scenario of a tumor that crosses the midline and an ipsilateral neck where at least LNL III is clinically involved, the risk for occult contralateral LNL II disease is around 8.5%. <!-- TODO: make this dynamic --> For the same scenario and an advanced T-category tumor, the risk increases to 11%. <!-- TODO: make this dynamic -->

Lastly, the predicted risk is also correlated via the time-steps to the degree of ipsilateral involvement. Changing the aforementioned scenario (advanced T-category, midline extension, ipsilateral LNL III clinically involved) to one where the patient presents with a clinically N0 neck, the risk for occult disease in the contralateral LNL II falls from 11% <!-- TODO: make this dynamic --> to 9.5%. <!-- TODO: make this dynamic -->

Taken together, T-category and ipsilateral involvement may still considerably impact the risk prediction for contralateral involvement: In @fig-model-risks the scenarios underlying the orange (6.5% <!-- TODO: make this dynamic -->) and the red (11% <!-- TODO: make this dynamic -->) histograms differ in T-category (early vs advanced) and clinical diagnosis (N0 vs ipsilateral at least LNL III involved).

@@

As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III may only cross a 5% threshold if the upstream level II is clinically involved, too. If the clinical diagnosis of the contralateral upstream level II is pathologically confirmed by FNA the risk increases by another percentage point<!-- TODO: make dynamic -->. When the tumor additionally crosses the mid-sagittal line, the risk may reach around 7%<!-- TODO: make dynamic -->.

@@

Our model's prediction for the contralateral LNL IV show that even in the most extreme case, with an advanced T-category tumor extending over the mid-sagittal line as well as metastases in every ipsilateral LNL and every contralateral upstream level, the risk stays just below 3%<!-- TODO: make dynamic -->. We only observe a risk of 5-6% when the involvement of the upstream level III is pathologically confirmed after taking a biopsy using FNA. This is because the model's prior probability for contralateral LNL IV involvement is so low that even given a clinical diagnosis in that level, a false positive in LNL III is a more likely explanation. This possibility, however, drops virtually to zero when tumor cells are pathologically confirmed in level III.

@@

- LNL II when the tumor extends over the mid-sagittal line
- LNL III when LNL II is clinically involved, may need confirmation via FNA
- LNL IV only when upstream level III pathologically confirmed

>>>>

@Fig-model-risks (left panel) shows the predicted risk of occult disease in contralateral LNL II. The most important variable impacting the prediction for contralateral level II involvement in our model is the tumor's lateralization. A patient with a clearly lateralized early T-category tumor and a clinically N0 neck is predicted to have a 1-2% <!-- TODO: make this dynamic --> risk for occult disease in contralateral LNL II. For a patient with a clearly lateralized early T-category tumor but *with* mid-sagittal extension, the risk increases to {{< var risk.early_ext_iN0_cN0_II >}}%.

Also advanced T-category increases the risk of occult disease but plays a lesser role: Considering the scenario of an early T-category tumor that crosses the midline and an ipsilateral neck where the LNLs II and III are clinically involved, the risk for occult contralateral LNL II disease is around {{< var risk.early_ext_iII+III_cN0_II >}}%. For the same scenario and an advanced T-category tumor, the risk increases to {{< var risk.late_ext_iII+III_cN0_II >}}%.

Lastly, the predicted risk is also correlated via the time-steps to the degree of ipsilateral involvement. When an early T-category patient with a tumor that extends over the mid-sagittal line presents with a clinically N0 ipsilateral neck instead of involvement in the levels II and III, the risk is {{< var risk.early_ext_iN0_cN0_II >}}% instead of {{< var risk.early_ext_iII+III_cN0_II >}}%.

Taken together, T-category and ipsilateral involvement may still considerably impact the risk prediction for contralateral involvement: In @fig-model-risks the scenarios underlying the orange ({{< var risk.early_ext_iN0_cN0_II >}}%) and the purple ({{< var risk.late_ext_iII+III_cN0_II >}}%) histograms differ in T-category (early vs advanced) and clinical diagnosis (N0 vs ipsilateral LNL II and III involved).

@@

As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III may only cross a 5% threshold if the upstream level II is clinically involved, too. If the clinical diagnosis of the contralateral upstream level II is pathologically confirmed by FNA the risk increases by another percentage point to {{< var risk.late_noext_iII_cII_FNA_III >}}%. When the tumor additionally crosses the mid-sagittal line, the risk may reach around {{< var risk.late_ext_iII_cII_FNA_III >}}%.

@@

Our model's prediction for the contralateral LNL IV show that even in the most extreme case, with an advanced T-category tumor extending over the mid-sagittal line as well as metastases in the ipsilateral LNLs II-IV and the contralateral upstream levels II and III, the risk stays at just {{< var risk.late_ext_iII+III+IV_cII+III_IV >}}%. We only observe a risk of {{< var risk.late_ext_iII+III+IV_cII+III_FNA_IV >}}% when the involvement of the upstream level III is pathologically confirmed after taking a biopsy using FNA. This is because the model's prior probability for contralateral LNL IV involvement is so low that even given a clinical diagnosis in that level, a false positive in LNL III is a more likely explanation. This possibility, however, drops virtually to zero when tumor cells are pathologically confirmed in level III.

## Contralateral LNLs I, V, and VII

Even for advanced T-category patients whose tumor crosses the midline and who present with extensive ipsi- and contralateral clinical involvement, the risk for involvement in the LNLs I and VII never exceeds 3%. LNL V contralaterally also shows only a very low risk for occult disease, but it does rise to {{< var risk.late_ext_iI+II+III+IV+V+VII_cII+III+IV_FNA_V >}}% in the same extreme case when additionally the upstream level IV involvement has been confirmed with FNA.

@@

- LNL II when the tumor extends over the mid-sagittal line
- LNL III when LNL II is clinically involved, may need confirmation via FNA
- LNL IV only when upstream level III pathologically confirmed
- LNLs I and VII only when clinically involved
- LNL V only in extreme cases (adv. T-cat., midext, all ipsi LNLs metastatic, contra II, III, and IV clinically involved plus cIV via FNA confirmed)
